### 2015-11-16 paper.Rmd
encounter rate variance update

Uncertainty in $\hat{N}$ comes from two sources:

@@

Sampling uncertainty can be characterised (rather than as simply $n$) by the *encounter rate*: $n/L$ for line transects (where $L$ is the total line length) or $n/K$ for point transects. Various formulations for the encounter rate variance are discussed in detail in @Fewster:2009ku. \pkg{Distance} implements all of the estimators of encounter rate variance given in that article plus some extras.

For line transects, by default, \pkg{Distance} uses a variation of the estimator "R2" from @Fewster:2009ku replacing number of observations per sample with the estimated abundance per sample (per @Innes:2002ka); that is for line transects:
$$
\widehat{\text{Var}}_{\text{encounter},R2} = \frac{K}{L^2(K-1)} \sum_{k=1}^{K} l_k^2 \left( \frac{N_k}{l_k} - \frac{N}{L}\right)^2,
$$
where $l_k$ are the lengths of the $K$ transects (such that $L = \sum_{k=1}^K l_k$).

For points,

The \code{varn} manual page gives further advice and technical detail on encounter rate variance.

>>>>

Here we take an intuitive approach to uncertainty estimation, for a full derivations consult @Marques:2003vb. Uncertainty in $\hat{N}$ comes from two sources:

@@

Sampling uncertainty can be characterised (rather than as simply $n$) by the *encounter rate*: the number of animals per unit transect. When covariates are not included in the detection function we can define the encounter rate as $n/L$ for line transects (where $L$ is the total line length) or $n/T$ for point transects (where $T$ is the total number of visits to all points). When covariates are included in the detection function, it is recommended that we substitute the $n$ in the previous expressions with the estimated abundance $\hat{N}$ as this will take into account the effects of the covariates.

For line transects, by default, \pkg{Distance} uses a variation of the estimator "R2" from @Fewster:2009ku replacing number of observations per sample with the estimated abundance per sample [@Innes:2002ka; @Marques:2003vb]:
$$
\widehat{\text{Var}}_{\text{encounter},R2} = \frac{K}{L^2(K-1)} \sum_{k=1}^{K} l_k^2 \left( \frac{N_k}{l_k} - \frac{N}{L}\right)^2,
$$
where $l_k$ are the lengths of the $K$ transects (such that $L = \sum_{k=1}^K l_k$). Whereas for points we use estimator "P3" from @Fewster:2009ku but again replacing $n$ by $\hat{N}$ to obtain the following estimator:
$$
\widehat{\text{Var}}_{\text{encounter},P3} = \frac{1}{T(K-1)} \sum_{k=1}^{K} t_k \left( \frac{N_k}{t_k} - \frac{N}{T}\right)^2,
$$
where $t_k$ is the number of visits to point $k$ and $T = \sum_{k=1}^K t_k$.

Various formulations for the encounter rate variance are discussed in detail in @Fewster:2009ku. \pkg{Distance} implements all of the estimators of encounter rate variance given in that article. The \code{varn} manual page gives further advice and technical detail on encounter rate variance.
